---
title: "Fisher's exact test failure can lead to biased results"
author: 
    - name: Robersy Sanchez
      affiliation: Department of Biology.
                   Pennsylvania State University, University Park, PA 16802
      email: rus547@psu.edu
date: "`r format(Sys.time(), '%d %B %Y')`"
fontsize: 11pt
fontfamily: "serif"
output:
    html_document:
        keep_md: yes
        toc: true
        toc_depth: 4
        toc_float: 
                collapsed: false
                smooth_scroll: true
        number_sections: true
        theme: united 
        geometry: margin=0.8in
        highlight: tango
toc_float: true
toc_depth: 4
abstract: |
    Fisher's exact test is a statistical significance test used in the analysis
    of contingency tables. Although this test is routinely used even though,
    it has been full of with controversy for over 80 years.Herein, the case of 
    its application analyzed is scrutinized with specific examples.
vignette: >
  %\VignetteIndexEntry{Fisher's exact test failure can lead to biased results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overwiew


The statistical significance of the difference between two bisulfite sequence
from control and treatment groups at each CG site can be evaluated with Fisher's
exact test. This is a statistical test used to determine if there are nonrandom
associations between two categorical variables.

Let there exist two such (categorical) variables $X$ and $Y$, where $X$ stands
for two groups of individuals: control and treatment, and $Y$ be a two states
variable denoting the methylation status, carrying the number of times that a
cytosine site is found methylated ($^{m}CG$) and non-methylated ($CG$),
respectively.

This information can be summarized in a $2 \times 2$ table, a $2 \times 2$
matrix in which the entries $a_{ij}$ represent the number of observations in
which $x=i$ and $y=j$. Calculate the row and column sums $R_i$ and $C_j$,
respectively, and the total sum:
  
$$N=\sum_iR_i=\sum_jC_j$$  
of the matrix:

|           | $Y = ^mCG$       | $Y = CG$         | $R_i$            |
|:----------|:-----------------|:-----------------|:-----------------|
| Control   | $a_{11}$         | $a_12$           | $a_{11}+a_{12}$  |
| Treatment | $a_{21}$         | $a_22$           | $a_{21}+a_{22}$  |
| $C_i$     | $a_{11}+a_{21}$  | $a_{12}+a_{22}$  | $a_{11}+a_{12}+a_{21}+a_{22} = N$ |

Then the conditional probability of getting the actual matrix, given the
particular row and column sums, is given by the formula:

$$P_{cutoff}=\frac{R_1!R_2!}{N!\prod_{i,j}a_{ij}!}C_1!C_2!$$

# Example 1

Let's consider the following hypthetical case of methylation at a given cytosine
site found in the comparison of control and treatment groups:
```{r examp1}
case1 <- matrix(c(5, 14, 15, 12),  nrow = 2, 
                     dimnames = list(Group = c("Ctrl", "Treat"),
                                     Meth.status = c("mCG", "CG")))
case1
```

That is, the cytosine site was found methylated 5 times from 20 counts in the
control group and 14 out of 26 in the treatment. This accounts for methylation
levels about 0.28 and 0.53 in the control and and the treatment groups,
respectively, which correspond to a value of 0.25 (50%) of methylation levels 
difference.
```{r pro1}
## Proportions
case1/rowSums(case1)
```

Fisher's exact test found not difference between these group!
```{r test1}
fisher.test(case1)
```

Considering the direction of the changes seems to be more sensitive to the
magnitude of methylation changes, statistically significant
at a significance level of $\alpha = 0.05$.
```{r test2}
fisher.test(case1, alternative = "less")
```


To realize how difficult would be the interpretation of this result in a real
concrete scenario, let's consider a basketball team where a top player finished
a game scoring 14 field-goal out of 26 shooting. Base on Fisher's exact test,
does it make sense to say that another player who made 6 out of 15 field-goals
performed as well as the best player?


### Bootstrap's tests for $2 \times 2$

Alternative testings are possible by means of bootstrap re-sampling from the set
(population) of all the matrices with the same row and column sums $R_i$ and
$C_j$. The analyses will be accomplished with function
[tableBoots](https://rdrr.io/github/genomaths/usefr/man/tableBoots.html) from
the R packge named [usefr](https://github.com/genomaths/usefr).

Hellinger test statistic has been proposed to compare discrete probability 
distributions ([1](#1)). Basu et all have shown that the Hellinger divergence,
as reported in [1](#1), has asymptotically has a chi-square distribution with
one degree of freedom under the null hypothesis. This a good property that makes
this statistic Hellinger statistic suitable for bootstrap's test of two discrete
populations. 
```{r boot1, message=FALSE}
library(usefr)
tableBoots(x = case1, stat = "hd", num.permut = 1999)
```


The Root-Mean-Square statistic has been also proposed to test differences 
between two discrete probability distributions ([2](#2))

```{r boot2, message=FALSE}
library(usefr)
tableBoots(x = case1, stat = "rmst", num.permut = 1999)

```


The $\chi^2$ statistic fails
```{r boot3, message=FALSE}
library(usefr)
tableBoots(x = case1, stat = "chisq", num.permut = 1999)
```

Assuming that expected discrete probability distribution (DPD) are:
```{r tp}
p <- case1[1,]/sum(case1[1,])
p
```

Then, we can test whether the treatment departs from the expected DPD:
```{r chisq}
chisq.test(x= case1[2,], p = p, 
           simulate.p.value = TRUE, B = 2e3)
```


## The example 1 is not unique

The failure of Fisher's exact test is evident in the comparisons given below:
```{r exampl2, message=FALSE}
res <- lapply(seq(1,78,1), function(k) {
    datos <- matrix(c(5, k+1, 15, k),  nrow = 2,
                    dimnames = list(Group = c("Ctrl", "Treat"),
                                    Meth.status = c("mCG", "CG")))
    datos
    x <- datos/rowSums(datos)
    x <- x[2,1] - x[1,1]
    ft <- fisher.test(datos)
    hd <- tableBoots(x = datos, stat = "hd", num.permut = 1999)
    rmst <- tableBoots(x = datos, stat = "rmst", num.permut = 1999)
    chisq <- tableBoots(x = datos, stat = "chisq", num.permut = 1999)

    
    chisq.p <- chisq.test(x= datos[2,], p = datos[1,]/sum(datos[1,]), 
                        simulate.p.value = TRUE, B = 2e3)$p.value
    
    c(meth.diff = x, 
      FT.p.value = ft$p.value,
      HD.p.value = hd,
      RMST = rmst,
      chisq = chisq,
      chisq.p = chisq.p)
})
do.call(rbind, res)
```

# Extreme situations found in methylation data sets

Unfortunate situations like that one shown below are not rare in methylation data.
The minimum coverage (total counts) in the example below is 5.
```{r examp3}
case2 <- matrix(c(3, 90, 2, 10),  nrow = 2, 
                     dimnames = list(Group = c("Ctrl", "Treat"),
                                     Meth.status = c("mCG", "CG")))
case2
```

The small numbers of count in the control sample yields a bias estimation of the
methylation level:

```{r pro2}
## Proportions
case2/rowSums(case2)
```

There is not enough experimental data to ensure that 3/(3+2) = 0.6 is an
unbiased estimated value of the methylation levels at the given cytosine site.
Fisher's exact test is insensitive to this experimental fact.

```{r test4}
fisher.test(case2)
```


```{r boot, message=FALSE}
set.seed(1)
st <- tableBoots(x = case2, stat = "hd", num.permut = 1999)
st
```



# References
1. <a name="1"></a> Basu A, Mandal A, Pardo L. Hypothesis testing for two
discrete populations based on the Hellinger distance. Stat Probab Lett. Elsevier
B.V.; 2010;80: 206–214. doi:10.1016/j.spl.2009.10.008.
2. <a name="2"></a> Perkins W, Tygert M, Ward R. Some deficiencies of χ2 and
classical exact tests of significance. Appl Comput Harmon Anal. Elsevier Inc.;
2014;36: 361–386. doi:10.1016/j.acha.2013.06.002

