---
title: "Non-linear Fit of Mixture Distributions"
author:
    - name: Robersy Sanchez
affiliation: Department of Biology.
             Pennsylvania State University, University Park, PA 16802
email: rus547@psu.edu
date: "`r format(Sys.time(), '%d %B %Y')`"
fontsize: 11pt
fontfamily: "serif"
output:
    html_document:
    keep_md: yes
toc: true
toc_depth: 4
toc_float:
    collapsed: false
smooth_scroll: true
number_sections: true
theme: united
geometry: margin=0.8in
highlight: tango
abstract: |
    Frequently, experimental data sets are forcely modeled by specific 
    probability distribution functions, when in fact they would be better 
    modelled by mixture of probability distributions. But, how to do it? 
    Where are the available tools? The _usefr_ R package provides some usefull
    tools to help the identification of the best mixture distribution model.
vignette: >
  %\VignetteIndexEntry{Fisher's exact test failure can lead to biased results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Background

It is said that a distribution $f(x)$ is a mixture $k$ distributions component:
$f_1(x), ..., f_k(x)$ if:


$$f(x) = \sum_{i=1}^k \pi_i f_i(x)$$

where $\pi_i$ are the so called mixing weights, $0 \le \pi_i \le 1$, and 
$\pi_1 + ... + \pi_k = 1$. More information about mixture distribution can be
read in [Wikipedia](https://en.wikipedia.org/wiki/Mixture_distribution).

Herein, we will show how to fit numerical data to a mixture of probability
distributions model using [usefr](https://genomaths.github.io/usefr/index.html)
R package. 

# Random values generated from a mixture of Gamma and Weibull distributions

To generate from a mixture distribution the function  
[mixtdistr](https://genomaths.github.io/usefr/reference/mixtdistr.html) will 
be used.

```{r lib, message=FALSE, warning=FALSE}
library(usefr) 

set.seed(3) # set a seed for random generation

# ========= A mixture of three distributions ========= 
phi = c(3/10, 7/10) # Mixture proportions 
# --------------------------------------------------------- 
# === Named vector of the corresponding distribution function parameters 
# must be provided 
args <- list(gamma = c(shape = 2, scale = 0.1), 
            weibull = c(shape = 3, scale = 0.5)) 
# ------------------------------------------------------------ 
# ===== Sampling from the specified mixture distribution ==== 
X <- rmixtdistr(n = 1e5, phi = phi , arg = args)
X <- na.omit(X)
X[1:20]
```

## The histogram of the mixture distribution

The graphics for the simulated dataset and the corresponding theoretical 
mixture distribution. The density curve is drawn applying function 
[dmixtdistr](https://genomaths.github.io/usefr/reference/mixtdistr.html):

```{r hist}
hist(X, 90, freq = FALSE, las = 1, family = "serif", 
panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
grid(col="white", lty = 1)}, family = "serif", col = "cyan1",
border = "deepskyblue", xlim = c(0, 1.5))
x1 <- seq(-4, 1.5, by = 0.001)
lines(x1, dmixtdistr(x1, phi = phi, arg = args), col = "red")
```


# Nonlinear fit of the mixture distribution

The nonlinear fit of this dataset is NOT straightforward! Herein, we used 
function [fitMixDist](https://genomaths.github.io/usefr/reference/fitMixDist.html)
```{r fit, error=TRUE}
FIT <- fitMixDist(X, args = list(gamma = c(shape = NULL, scale = NULL),
                                weibull = c(shape = NULL, scale = NULL)), 
                npoints = 50, usepoints = 1000)
summary(FIT$fit)
```

It seems to be that, depending on the OS, numerical algorithm variant, or the 
computer processor, the Cholesky factorization of a real symmetric 
positive-definite square matrix can fail (see for example: the same example at:
[genomaths.com](https://genomaths.com/stats/non-linear-fit-of-mixture-distributions/)).

Fortunately, to bypass the numerical issue, we can apply a numerical trick.
In the current case, Gamma and Weibull distribution of this data are not quite
different. So, firstly, we can try the fitting to a mixture of Weibull 
distributions and to get some guessing for starting parameter values for each 
distribution.

```{r fit-1}
FIT <- fitMixDist(X, args = list(weibull = c(shape = NULL, scale = NULL),
                                weibull = c(shape = NULL, scale = NULL)), 
                npoints = 50, usepoints = 1000)
summary(FIT$fit)
```

Next, the estimated parameter values are supplied as starting parameter values
for the mixture of Gamma and Weibull distributions, and the numerical problem 
is solved:

```{r fit-2}
FIT <- fitMixDist(X, args = list(gamma = c(shape = 1.55, scale = 0.31),
                                weibull = c(shape = 3.4, scale = 0.52)), 
                npoints = 50, usepoints = 1000)
summary(FIT$fit)
```

## Graphics of the simulated dataset and the corresponding theoretical mixture distribution

Function [dmixtdistr](https://genomaths.github.io/usefr/reference/mixtdistr.html)
is applied to draw the density curve:
```{r den-plot, fig.height=6}
hist(X, 90, freq = FALSE, las = 1, family = "serif", 
    panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
                grid(col="white", lty = 1)},
    family = "serif", col = "seagreen1", border = "deepskyblue", 
    xlim = c(0, 1.5), cex.lab = 1.2, main = "")

x1 <- seq(-4, 10, by = 0.001)
lines(x1, dmixtdistr(x1, phi = FIT$phi, arg = FIT$args), col = "red")
mtext("Histogram of Gamma & Weibull Mixture Distributions.",
    cex = 1.4, font = 3, family = "serif")
```

## The GOF test for the mixture distribution

## Bootstrap goodness-of-fit test

A bootstrap goodness-of-fit (GOF) test is performed with function
[mcgoftest](https://genomaths.github.io/usefr/reference/mcgoftest.html).

The parameter values are taken from the previous fitted mixture distribution
model. Notice the particular way to set up the list of parameters. The Null
hypothesis is that the data set follows a mixture of Gamma and Weibull
distributions with the estimated parameter values.

```{r mcgoftest}
pars <- c(list(phi = FIT$phi), arg = list(FIT$args))
mcgoftest(varobj = X, distr = "mixtdistr", pars = pars, 
        num.sampl = 999, sample.size = 99999, stat = "chisq", 
        num.cores = 4, breaks = 200, seed = 123)
```


# Comparison with a forced single distribution model

```{r hist2}
set.seed(3) # set a seed for random generation

# ========= A mixture of three distributions ========= 
phi = c(3/10, 7/10) # Mixture proportions 

# ------------------------------------------------------------ 
# === Named vector of the corresponding distribution function parameters 
# must be provided 
args <- list(gamma = c(shape = 2, scale = 0.1), 
            weibull = c(shape = 2.1, scale = 0.12)) 
# ------------------------------------------------------------ 

# ===== Sampling from the specified mixture distribution ==== 
X <- rmixtdistr(n = 1e5, phi = phi , arg = args)

# ---------------------- Histogram ---------------------------

hist(X, 90, freq = FALSE, las = 1, family = "serif", 
    panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
                grid(col="white", lty = 1)},
    family = "serif", col = "seagreen1", border = "deepskyblue", 
    xlim = c(0, 1.5), cex.lab = 1.2, main = "")

x1 <- seq(-4, 10, by = 0.001)
lines(x1, dmixtdistr(x1, phi = phi, arg = args), col = "red")
mtext("Histogram of Gamma & Weibull Mixture Distributions.",
    cex = 1.4, font = 3, family = "serif")
```


## The single distribution model
Next, a search of the best single probability distribution model is accomplished
with function [fitCDF](https://genomaths.github.io/usefr/reference/fitCDF.html).

The generalized Gamma distribution model seems to be the best fitted model,
followed by the Gamma model.
```{r fit2, warning=FALSE, fig.height=8}
cdfp <- fitCDF(X,
    distNames = c("Gamma", "Weibull", "3P Weibull", "Generalized 3P Gamma"),
    xlabel = "My Nice Variable Label",
    plot = TRUE, plot.num = 2, font.lab = 3, font = 2, 
    font.axis = 2, family = "serif", cex.text = 1.2,
    cex.main = 1.4, cex.lab = 1.4, cex.axis = 1.4
)
```


Clearly, without the previous knowledge about the true CDF, the graphics cast a
suspicion about how good the best fitted model is. Althout the Q-Q plot is not
good, the cross-validation correlation coefficient R (R.Cross.val) support a
strong average of cross-validation predictive power of the generalized Gamma
distribution model:
```{r crosv}
cdfp$gof
```

Also, the GoF testing based on the Monte Carlos Anderson–Darling statistic does
not reject the model.

```{r mcgoftest1}
mcgoftest(varobj = X, 
        distr = "ggamma", 
        pars = cdfp$bestfit$par, 
        num.sampl = 999, stat = "ks",
        sample.size = 99999, 
        num.cores = 4)
```

## The mixture distribution model

As before the Weibull distribution model seems to be the best mixture model
```{r fit-4}
FIT1 <- fitMixDist(X, args = list(weibull = c(shape = NULL, scale = NULL),
                                weibull = c(shape = NULL, scale = NULL)), 
                npoints = 100, usepoints = 2000)
summary(FIT$fit)
```

The mixture distribution model using the Gamma as one of marginal distributions
fails. 
```{r fit5, error=TRUE}
FIT2 <- fitMixDist(X, args = list(gamma = c(shape = NULL, scale = NULL),
                                weibull = c(shape = NULL, scale = NULL)), 
                npoints = 100, usepoints = 2000)
summary(FIT2$fit)
```

Next, the estimated parameter values from the Weibull mixture distribution model
are applied as starting parameter values to get the best fitting model for
a Gamma-Weibull mixture distribution model.
```{r fit4}
FIT2 <- fitMixDist(X, args = list(gamma = FIT1$fit$par[1:2],
                                weibull = FIT1$fit$par[1:2]), 
                npoints = 100, usepoints = 2000)
summary(FIT2$fit)
```

## The Q-Q Plot of the alternative models

```{r qqplot, fig.width=12, fig.height=8}
par(mfcol = c(2, 2))
xl <- c(min(X, na.rm = TRUE), max(X, na.rm = TRUE))
n = 1e5

### ---------------------------------------------------------------------
pars <- cdfp$bestfit$par
set.seed(3) # set a seed for random generation
xl <- c(min(X, na.rm = TRUE), max(X, na.rm = TRUE))

q = rggamma(n = n, alpha = pars[1], scale = pars[2], psi = pars[3])

qqplot(x = q, y = X, las = 1, family = "serif", bty="n",
    panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
                grid(col="white", lty = 1)},
    family = "serif", col = "red",  cex.lab = 1.2, 
    main = expression("Q-Q plot for the Generalized Gamma distribution model"),
    xlab = "Theoretical Quantiles", ylab = "Sample Quantiles",
    xlim = xl, ylim = xl)
qqline(y = X, 
    distribution = function(p) 
        qggamma(p, alpha = pars[1], scale = pars[2], psi = pars[3]),
    probs = c(0.1, 0.6), col = 2)

### ---------------------------------------------------------------------


q = rmixtdistr(n = n, phi = FIT1$phi , arg = FIT1$args)
qqplot(x = q, y = X, las = 1, family = "serif", bty="n",
    panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
                grid(col="white", lty = 1)},
    family = "serif", col = "red",  cex.lab = 1.2, 
    main = expression("Q-Q plot for the Weibull mixture distribution model"),
    xlab = "Theoretical Quantiles", ylab = "Sample Quantiles",
    xlim = xl, ylim = xl)
qqline(y = X, 
    distribution = function(p) qmixtdistr(p, phi = FIT1$phi , arg = FIT1$args),
    probs = c(0.1, 0.6), col = 2)

### ---------------------------------------------------------------------

set.seed(3) # set a seed for random generation
pars = cdfp$fit$Gamma$par

q = rgamma(n = n, shape = pars[1], scale = pars[2])

qqplot(x = q, y = X, las = 1, family = "serif", bty="n",
    panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
                grid(col="white", lty = 1)},
    family = "serif", col = "red",  cex.lab = 1.2, 
    main = expression("Q-Q plot for the Gamma distribution model"),
    xlab = "Theoretical Quantiles", ylab = "Sample Quantiles",
    xlim = xl, ylim = xl)
qqline(y = X, 
    distribution = function(p) 
        qgamma(p, shape = pars[1], scale = pars[2]),
    probs = c(0.1, 0.6), col = 2)

### ---------------------------------------------------------------------

q = rmixtdistr(n = n, phi = FIT2$phi , arg = FIT2$args)
qqplot(x = q, y = X, las = 1, family = "serif", bty="n",
    panel.first={points(0, 0, pch=16, cex=1e6, col="grey95")
                grid(col="white", lty = 1)},
    family = "serif", col = "red", cex.lab = 1.2, 
    main = expression("Q-Q plot for the Gamma-Weibull mixture distribution model"),
    xlab = "Theoretical Quantiles", ylab = "Sample Quantiles",
    xlim = xl, ylim = xl)
qqline(y = X, 
    distribution = function(p) qmixtdistr(p, phi = FIT2$phi , arg = FIT2$args),
    probs = c(0.1, 0.6), col = 2)
```

The Q-Q plots suggest that the Gamma-Weibull mixture distribution model is the
best fitted model.

## The GOF test for the mixture distribution

The Kolmogorov-Smirno test does not reject the Null hypothesis about that the 
_X_ follows Gamma-Weibull mixture distribution model. 
```{r gof_mixture}
set.seed(1)
pars <- c(list(phi = FIT2$phi), arg = list(FIT2$args))
mcgoftest(varobj = X, distr = "mixtdistr", pars = pars, 
        num.sampl = 999,
        sample.size = 99999, stat = "ks", 
        num.cores = 4, seed = 123)
```


Although the Monte Carlo $mc_p.value = 0.316$ is lower than the value estimated
for the generalized gamma model ($mc_p.value = 1$), the 
$KS.stat.p.value = 0.289$ of the simple Kolmogorov-Smirno test support the
mixture model, which for case of the generalized gamma model 
($KS.stat.p.value = 0$) is reject.


# Conclusions


To identify the best fitted probability distribution model for an experimental
data set would be a laborious task. Usually, researcher move forward with the
first single CDF model that is supported by the Goodness-of-Fit tests. However,
digging deep in the structure of the experimental data, it would turn out that
the data would come from a stratified population, better modelled by mixture of
probability distribution functions.


# Session info {.unnumbered}

Here is the output of `sessionInfo()` on the system on which this document was
compiled running pandoc `r rmarkdown::pandoc_version()`:


```{r sessionInfo, echo=FALSE}
sessionInfo()
```






