---
title: "Get started-with 'usefr'"
author: 
    - name: Robersy Sanchez
      affiliation: Department of Biology.
                   Pennsylvania State University, University Park, PA 16802
      email: rus547@psu.edu
date: "`r format(Sys.time(), '%d %B %Y')`"
fontsize: 11pt
fontfamily: "serif"
output:
    html_document:
    keep_md: yes
toc: true
toc_depth: 4
toc_float:
    collapsed: false
smooth_scroll: true
number_sections: true
theme: united
geometry: margin=0.8in
highlight: tango
abstract: |
    Package 'usefr' is collection of functions for statistical analysis of
    nonlinear trends and corresponding goodness-of-fit test. Herein, a fast
    starting introduction is provided.
vignette: >
  %\VignetteIndexEntry{"Get started-with 'usefr'"}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Install and loading the R packages needed for the computation

*Install R dependencies:*      
```{r install1, eval=FALSE}
    if (!requireNamespace("BiocManager")) install.packages("BiocManager")
    BiocManager::install()
    
    BiocManager::install(c("BiocParallel","minpack.lm", "numDeriv", "copula", 
                            "mclust", "nls2", "cubature", "mixdist"), 
                        dependencies=TRUE)
```

*Install 'userfr':*       
```{r install2, eval=FALSE}
BiocManager::install("genomaths/usefr")
```

Load the libraries needed:
```{r lib, warning=FALSE, message=FALSE}
library(usefr)
library(ggplot2)
library(gridExtra)
```


# Nonlinear fit of a commutative distribution functions

Usually the parameter estimation of a cumulative distribution
function (*CDF*) are accomplished using the corresponding probability density
function (*PDF*). Different optimization algorithms can be used to
accomplished this task and different algorithms can yield different estimated
parameters. However, the *CDF* is required to accomplish any useful prediction 
of the probability of an event described by *PDF*. Hence, to bypass the 
intermediate numerical error, why not try to fit the CDF directly?

Let us simulate random samples from a specified Normal and Weibull
distributions. To make reproducible this example, we set a seed.
```{r sim1}
set.seed(1)
## Parameters
mean = 1.5
sd = 2
shape = 3.1
scale = 4
mu = 5

## Simulating samples
x1 = rnorm(10000, mean = mean, sd = sd)
x2 = rweibull3p(10000, shape = shape, scale = scale, mu = mu)
dt <- data.frame(x1 = x1, x2 = x2)
dt[1:10,]
```

## The histograms and density plots of the given Normal and Weibull distributions

```{r hist, fig.width=13, fig.height=6}
p1 <- ggplot(data = dt, aes(x1)) + 
  geom_histogram(data = dt, aes(y=after_stat(density)), binwidth = 1, 
                 colour = "black", fill = "skyblue", na.rm=TRUE) +
  stat_function(fun = dnorm, n = 101, col = "red",
                args = list(mean = mean, sd = sd), linewidth = 1) +
  theme_gray(base_family = "serif", base_size = 14) +
  annotate(geom = "text", x = 7, y = 0.16, size = 6,
           label = 'bolditalic(N(1.5,2))',
           family = "serif", color = "blue", parse = TRUE)  

p2 <- ggplot(data = dt, aes(x2)) + 
  geom_histogram(data = dt, aes(y=after_stat(density)), binwidth = 1, 
                 colour = "black", fill = "skyblue", na.rm=TRUE) + 
  #xlim(0,20) + ylim(0,0.23) +
  stat_function(fun = dweibull3p, n = 101, col = "red",
                args = list(shape = shape, scale = scale, mu = mu), 
                linewidth = 1) +
  theme_gray(base_family = "serif", base_size = 14) +
  annotate(geom = "text", x = 12, y = 0.23, size = 6,
           label = 'bolditalic(W(3.1, 4, 5))',
           family = "serif", color = "blue", parse = TRUE)
grid.arrange(p1, p2, nrow = 1)
```

Notice that the above drawn densities are, in fact, empirical density 
estimations, based on in algorithmic kernel density estimation approaches. 
A typical situation found with experimental data sets is that frequently
we assume that sample is taken from a population that follows normal 
distribution when in fact the population follows a Weibull distribution. 
Indeed, a sample of a variable following Weibull distribution can fit normal
distribution model.

## Nonlinear fit to the Gaussian model
After the application of function fitCDF, results suggest that variable *x1*, 
sampled from a population following normal distribution, fits the normal 
distribution model:
```{r fit, warning=FALSE}
cdfp <- fitCDF(dt$x1,
    distNames = c("Normal", "Weibull"), xlabel = "My Nice Variable Label",
    plot = T, font.lab = 3, font = 2, font.axis = 2, family = "serif",
    cex.lab = 1.3, cex.axis = 1.3
)
```


The residual sum-of-squares is lower for the normal distribution model:
```{r sq}
cdfp$fit
```


### The Monte Carlo GOF test

The application of a Monte Carlo GOF test, accomplished with function 
mcgoftest, support the results of the previous nonlinear regression model:
```{r gof}
mcgoftest(varobj = dt$x1, distr = "norm", pars = c(1.5, 2), num.sampl = 500,
          sample.size = 1000, num.cores = 1)
```

```{r gof}
mcgoftest(varobj = dt$x1, distr = "weibull", pars = c(1.5, 2), num.sampl = 500,
          sample.size = 1000, num.cores = 1)
```

## Nonlinear fit to the Weibull model

```{r fit, warning=FALSE, fig.height=8}
cdfp <- fitCDF(dt$x2,
    distNames = c("Weibull", "3P Weibull", "Normal"),
    xlabel = "My Nice Variable Label",
    plot = TRUE, plot.num = 2, font.lab = 3, font = 2, 
    font.axis = 2, family = "serif", cex.text = 1.2,
    cex.main = 1.4, cex.lab = 1.4, cex.axis = 1.4
)
```

The residual sum-of-squares is lower for the 3-parameter Weibull distribution
model. However, the normal distribution model is also plausible:
```{r sq}
cdfp$fit
```

The Akaike information criteria suggest that the Weibull probability model is 
the best fitted model, which is confirmed (by a narrow margin) and by the 
cross-validation correlation coefficient R (R.Cross.val).
```{r}
data.frame(
    Weibull3P = cdf_crossval(
    model = cdfp$bestfit,  q = dt$x2),
    'AIC-Weibull3P' = cdfp$aic$AIC[1],
    Normal = cdf_crossval(
    model = cdfp$fit$Normal,  q = dt$x2),
    'AIC-Normal' = cdfp$aic$AIC[2],
    row.names = NULL
)
```

# The Monte Carlo GOF test
The Monte Carlo GOF test is accomplished with function mcgoftest.

Let's take the parameter values for the normal distribution model of variable
*x2*, which was sample from a population following a Weibull distribution.
```{r norm}
summary(cdfp$fit$Normal)
```


The application of a Monte Carlo GOF test, accomplished with function 
mcgoftest, support the modeling of variable *x2* with a normal distribution
model:
```{r gof}
mcgoftest(varobj = dt$x2, distr = "norm", pars = c(8.6, 1.3), num.sampl = 500,
          sample.size = 1000, num.cores = 1)
```

Both GOF, the Monte Carlo and classical Kolmogorov-Smirnov (KS), did not rejected 
normal distribution model. In fact, the KS statistics are quite close.

```{r gof1}
mcgoftest(varobj = dt$x2, distr = "weibull3p", 
          pars = c(shape = shape, scale = scale, mu = mu), 
          num.sampl = 500, sample.size = 1000, num.cores = 1)
```


The Anderson–Darling statistic reject the normal distribution model of variable
*x2*

```{r gof2}
mcgoftest(varobj = dt$x2, distr = "norm", pars = c(8.6, 1.3), 
          num.sampl = 500, stat = "ad",
          sample.size = 1000, num.cores = 1)
```


```{r gof3}
mcgoftest(varobj = dt$x2, distr = "weibull3p", stat = "ad",
          pars = c(shape = shape, scale = scale, mu = mu), 
          num.sampl = 500, sample.size = 1000, num.cores = 1)
```


Users can read the available tutorials see applications, e.g., on mixture
distribution models.


